The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
intro.

%\emph{Simultaneously} 
The execution begins at line~(0) with the value {{20}}. At line~(1)
values {{20}} and {{()}} flow forward, corresponding to the type
`{{a*1}}'. At line~(2) three independent subcomputations \emph{proceed
  concurrently}: the intial value {{20}} flowing from line (1) and two
entangled values, let us call them {{1/x}} and {{x}}, such that
specifying one implicitly specifies the dependant other. These three
subcomputations exist \emph{at the same time}. At line (3), the
dependant {{1/x}} is fixed to be the value {{20}}, thereby making
fixing values of all three branches.
%% the positive value created at line (2) which can flow to the
%% consumer; the negative value created at line (3) which can pay for the
%% debt created at line (2); and the initial value from line (0) which
%% can flow to the required forward-flowing value needed at line (3). 
At some intuitive level, one might understand the situation as
modeling a credit card payment in which money is created at line (2)
for the consumer together with a corresponding debt. The producer's
bank pays for the debt at line (3) and waits for the actual money to
arrive from the producer.

There are at least two fundamental points about this example that must be
emphasized:
\begin{itemize}
\item As the example illustrates, negative types corresponds to
  ``debts'' satified by backtracking and fractional types correspond
  to debts satisfied by constraint propagation.  It would clearly be
  disastrous if debts could be deleted or duplicated. This simple
  observation explains why these types are much simpler and much more
  appealing in a framework where information is guaranteed to be
  preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed
  to constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless
  in its treatment of information.
\item The appeal of negative and fractional types can already be
  gleaned from the credit card analogy. The main reason credit card
  transactions are convenient is because they disentangle the
  propagation of the resources (money) from the propagation of the
  services. Not every transaction needs both the resources and
  services to be brought together: it is sufficient to have a promise
  that the demand for resources will be somehow satisfied, as long as
  the infrastructure can be trusted with such
  promises. 
%% Computationally, this means that negative types allow us
%%   to break the dependency between producers and consumers and let each
%%   compute independently.

@@

\paragraph*{Functions and Dependancy Splitting.} 
Consider the following realization of boolean negation written using a
pattern-matching style:
\begin{verbatim}
notB b = case b of 
  False -> True
  True -> False
\end{verbatim}
In a functional language, we are not used to treating the pattern-matching
clauses \verb|False -> True| and \verb|True -> False| as first-class entities
but fractional types allow us to do just that! Informally, we can treat each
of the pattern-matching clauses as a first-class value of type 
{{(1/bool) * bool}}. 
A small but complete example will illustrate the main aspects of this idea.

Let \verb|color| be the datatype with three constructors \verb|R|, \verb|G|,
and \verb|B| and let's try to construct a function that rotates the
colors. In other words, we would like to build the following three
pattern-matching clauses \verb|R -> G|, \verb|G -> B|, and \verb|B -> R|.  In
the algebra of rational numbers we have that for each {{a}} that is not equal
to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic identity
corresponds to a type isomorphism {{1 <-> (1/a) * a}}. For the datatype
\verb|color|, there are three possible witnesses for this isomorphism mapping
{{()}} to either {{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the
particular witness determined by further constraints encountered during the
computation. The idea of variables that capture constraints that are
satisfied later is at the heart of logic programming and hence the witness for
the isomorphisms above is generally {{(1/alpha,alpha)}} where~{{alpha}} is a
fresh logic variable.

Returning to our goal of creating the function that rotates the colors, we
proceed as follows:
\begin{itemize}
\item Using the fact that {{1}} is the multiplicative unit, generate from the
  input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
  components of the above tuple. The resulting value is 
{{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
{{alpha2}}, and {{alpha3}} are fresh logic variables.
\item Using the fact that {{*}} is associative and commutative, we can
  rearrange the above tuple to produce the value:

{{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
\item At this point, we have produced the skeleton for our desired function.
  Assume now that a client uses the first clause of the function in a context
  that contains the value \verb|R|, i.e., the client has the value
  {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
  and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
  At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
  the first component of the pair which has the following effects:
  \begin{itemize}
    \item the logic variable {{alpha1}} is unified with {{R}};
    \item the first component of the pair then reduces to () and can be
      absorbed by the fact that {{1}} is the multiplicative unit. In other
      words, the value reduces to simply {{alpha2}} reflecting that the
      pattern-matching was successful and the result of the clause is now
      available;
    \item the first clause of the function has now disappeared and the
      remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
  \item The same client or another client can now use the second clause
    applying it to {{Gr}} which would produce the result {{alpha3}} for the
    second client and ground the result of the first client to {{Gr}}.
  \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
    which grounds the result of the second client to {{B}}.
  \end{itemize}
\end{itemize}
It should be evident that in a universe in which information is not
guaranteed to be preserved by the computational infrastructure, the above
slicing and dicing of functions as a collection of first-class pattern
clauses would make no sense. However, in a computational infrastructure that
is guaranteed to preserve information, computations such as the one above are
guaranteed to ``make sense.'' If one branch uses a clause that matches one
part of the input to produce one part of the output, all other clauses are
automatically adjusted to account for this to ensure that this input is not
matched again and that this output is not produced again. Furthermore, just
like the case for negative types, a dual reading of the pattern-matching
clauses as continuations allows the context to use the result of the clause
and provide the matching argument later.

>>>>

%\emph{Simultaneously} 
The execution begins at line~(0) with the value {{20}}. At line~(1)
values {{20}} and {{()}} flow forward, corresponding to the type
`{{a*1}}'. At line~(2) three independent subcomputations \emph{proceed
  concurrently}: the intial value {{20}} flowing from line (1) and two
entangled values, let us call them {{1/x}} and {{x}}, such that
specifying one implicitly specifies the dependant other. These three
subcomputations \emph{exist at the same time}. At line (3), the
dependant {{1/x}} is fixed to be the value {{20}}, thereby making
fixing values of all three branches.
%% the positive value created at line (2) which can flow to the
%% consumer; the negative value created at line (3) which can pay for the
%% debt created at line (2); and the initial value from line (0) which
%% can flow to the required forward-flowing value needed at line (3). 
At some intuitive level, one might understand the situation as
modeling a credit card payment in which money is created at line (2)
for the consumer together with a corresponding debt. The producer's
bank pays for the debt at line (3) and waits for the actual money to
arrive from the producer.  There are at least two fundamental points
about this example that must be emphasized:
\begin{itemize}
\item As the example illustrates, negative types corresponds to
  ``debts'' satified by backtracking and fractional types correspond
  to debts satisfied by constraint propagation.  It would clearly be
  disastrous if debts could be deleted or duplicated. This simple
  observation explains why these types are much simpler and much more
  appealing in a framework where information is guaranteed to be
  preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed
  to constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless
  in its treatment of information.
\item The appeal of negative and fractional types can already be
  gleaned from the credit card analogy. The main reason credit card
  transactions are convenient is because they disentangle the
  propagation of the resources (money) from the propagation of the
  services. Not every transaction needs both the resources and
  services to be brought together: it is sufficient to have a promise
  that the demand for resources will be somehow satisfied, as long as
  the infrastructure can be trusted with such
  promises. 
%% Computationally, this means that negative types allow us
%%   to break the dependency between producers and consumers and let each
%%   compute independently.

@@

\paragraph*{Functions and Dependancy Splitting.} 
%% Consider the following realization of boolean negation written using a
%% pattern-matching style:
%% \begin{verbatim}
%% notB b = case b of 
%%   False -> True
%%   True -> False
%% \end{verbatim}
%% In a functional language, we are not used to treating the pattern-matching
%% clauses \verb|False -> True| and \verb|True -> False| as first-class entities
%% but fractional types allow us to do just that! Informally, we can treat each
%% of the pattern-matching clauses as a first-class value of type 
%% {{(1/bool) * bool}}. 
%% A small but complete example will illustrate the main aspects of this idea.

%% Let \verb|color| be the datatype with three constructors \verb|R|, \verb|G|,
%% and \verb|B| and let's try to construct a function that rotates the
%% colors. In other words, we would like to build the following three
%% pattern-matching clauses \verb|R -> G|, \verb|G -> B|, and \verb|B -> R|.

In the algebra of rational numbers we have that for each {{a}} that is
not equal to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic
identity corresponds to a type isomorphism {{1 <-> (1/a) * a}}, whose
witness maybe thought of as the `identity function' at {{a}}.  For
example, for the datatype \verb#color = R|G|B#, there are three
possible witnesses for this isomorphism mapping {{()}} to either
{{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the particular
witness determined by further constraints encountered during the
computation.  The idea of variables that capture constraints that are
satisfied later is at the heart of logic programming and hence the
witness for the isomorphisms above is generally {{(1/alpha,alpha)}}
where~{{alpha}} is a fresh logic variable.

%% Returning to our goal of creating the function that rotates the colors, we
%% proceed as follows:

\begin{itemize}
\item Using the fact that {{1}} is the multiplicative unit, generate
  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
  the components of the above tuple. The resulting value is
  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}}, and
  {{alpha2}} are fresh logic variables.
\item Using the fact that {{*}} is associative and commutative, we can
  rearrange the above tuple to produce the value:

{{((1/alpha1,alpha2),(1/alpha2,alpha1))}}.

%% \begin{itemize}
%% \item Using the fact that {{1}} is the multiplicative unit, generate from the
%%   input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
%% \item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
%%   components of the above tuple. The resulting value is 
%% {{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
%% {{alpha2}}, and {{alpha3}} are fresh logic variables.
%% \item Using the fact that {{*}} is associative and commutative, we can
%%   rearrange the above tuple to produce the value:

%% {{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.

%% \item At this point, we have produced the skeleton for our desired function.
%%   Assume now that a client uses the first clause of the function in a context
%%   that contains the value \verb|R|, i.e., the client has the value
%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
%%   the first component of the pair which has the following effects:
%%   \begin{itemize}
%%     \item the logic variable {{alpha1}} is unified with {{R}};
%%     \item the first component of the pair then reduces to () and can be
%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
%%       words, the value reduces to simply {{alpha2}} reflecting that the
%%       pattern-matching was successful and the result of the clause is now
%%       available;
%%     \item the first clause of the function has now disappeared and the
%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
%%   \item The same client or another client can now use the second clause
%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
%%     second client and ground the result of the first client to {{Gr}}.
%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
%%     which grounds the result of the second client to {{B}}.
%%   \end{itemize}
\end{itemize}

At this point we have constructed a strange mixing of two functions;
`inputs' of one form `outputs' of the other. If {{(1/alpha1,alpha2)}}
is held by one subcomputation and {{1/alpha2,alpha1)}} is held by
another subcomputation, these `remixed functions' form a communication
channel between the subcomputations -- unifying {{1/alpha1}} with
{{R}} in one subcomputation fixes it in the other. Depending on the
choice of the type {{a}}, an arbitrary number of bits may be
communicated. The dual reading of the above, with additive functions
of the form {{0<-> ((-a)+b)}}, correspond to control flow and
corotuining between two subcomputations -- only one of which `exists'
at any point.

It should be evident that in a universe in which information is not
guaranteed to be preserved by the computational infrastructure, the
above slicing and dicing of functions would make no sense. However, in
a computational infrastructure that is guaranteed to preserve
information, computations such as the one above are guaranteed to
``make sense.''

%% If one branch uses a clause that matches one part of the input to
%% produce one part of the output, all other clauses are automatically
%% adjusted to account for this to ensure that this input is not matched
%% again and that this output is not produced again.

%% \item At this point, we have produced the skeleton for our desired function.
%%   Assume now that a client uses the first clause of the function in a context
%%   that contains the value \verb|R|, i.e., the client has the value
%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
%%   the first component of the pair which has the following effects:
%%   \begin{itemize}
%%     \item the logic variable {{alpha1}} is unified with {{R}};
%%     \item the first component of the pair then reduces to () and can be
%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
%%       words, the value reduces to simply {{alpha2}} reflecting that the
%%       pattern-matching was successful and the result of the clause is now
%%       available;
%%     \item the first clause of the function has now disappeared and the
%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
%%   \item The same client or another client can now use the second clause
%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
%%     second client and ground the result of the first client to {{Gr}}.
%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
%%     which grounds the result of the second client to {{B}}.
%%   \end{itemize}

%% Furthermore, just like the case for negative types, a dual reading of
%% the pattern-matching clauses as continuations allows the context to
%% use the result of the clause and provide the matching argument later.
